Embedding Intuitionistic into Classical Logic

### 2022-10-13 FoSSaCS/fossacs-EIICL.tex
asdf

The interest in intuitionistic logic has lead to the development of a number of automated theorem proving systems and a collection of benchmark problems (see e.g. the ILTP library website~\cite{iltp}).
However, intuitionistic provers perform poorly compared to their classical counterparts, see e.g. the TPTP~\cite{casc} and SAT~\cite{satc} competitions.
%, e.g., the first-order provers~\cite{kovacs2013first, schulz2002brainiac, korovin2008iprover}
%Comparing benchmarks for classical and intuitionistic automated theorem proving, e.g. TPTP~\cite{casc} and , we see that classical provers currently have a far better success rate. Furthermore all the most popular first-order utilize classical logic.
This difference can be partially explained by fundamental differences between the logics.
First of all, determining intuitionistic validity is computationally harder, i.e. in the propositional case intuitionistic validity is \verb+PSPACE+-complete~\cite{statman1979intuitionistic} whereas classical validity is \verb+coNP+-complete~\cite{cook1971complexity}.
A further advantage of classical logic is the existence of calculi that are particularly suited for automation such as superposition~\cite{bachmair2001resolution}, which rely on the existence of convenient normal forms such as CNF, and the duality between validity and satisfiability, i.e. to show the validity of a formula it suffices to show the unsatisfiability of the negated formula, which is insufficient in intuitionistic logic.
The first dedicated intuitionistic theorem provers~\cite{mclaughlin2009efficient,tammet1996resolution} used the naive inverse method, a direct search for a cut-free proof by applying the rules from some proof calculus inversely, which generally leads to a very complex search. More recently connection-based methods have been applied to various non-classical logics~\cite{otten2005clausal,otten2021nanocop}, including intuitionistic logic.
There have also been some successful attempts to study intuitionistic validity via embedding into higher-order classical logic~\cite{LEO}.
Finally, we add that in contrast to intuitionistic provers a tremendous amount of work has been put into optimizing provers for classical logic.

With this work we want to examine a new but evident approach for directly leveraging the progress in classical reasoning for intuitionistic logic.
To this end, we give an embedding from intuitionistic logic into classical logic. That is, we give an effective procedure that returns for each formula $\varphi$ a modified formula $\varphi^\#$, such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid.
Interestingly, the reverse direction, the famous double-negation translation, has long been established and goes back to Glivenko~\cite{glivenko1929quelques} in the propositional case, and to G\"odel~\cite{godel1933intuitionistischen} and Gentzen~\cite{gentzen1936widerspruchsfreiheit} in the first-order case. In the propositional case it is particularly simple: $\varphi$ is classically valid if and only if $\neg\neg\varphi$ is intuitionistically valid. Intuitively the translation collapses for each subformula $\psi$ of $\varphi$ the truth values of $\psi$ and $\neg\neg\psi$, which are classically but not intuitionistically equivalent. This gives us a first idea why the reverse direction is perhaps more difficult: We need to expand the truth values of $\psi$ and $\neg\neg\psi$, i.e. if they both occur in $\varphi$, we must have a way to (classically) assign different truth values to their respective counterparts in $\varphi^\#$. In particular, this necessitates the introduction of new propositional variables in our procedure, which already marks a big difference to the double negation translation.
%
In addition to the translation of formulas we give an effective translation of counter-models, i.e. for each intuitionistic counter-model of $\varphi$ we can effectively construct a classical counter-model of $\varphi^\#$ and vice versa.
We note that the existence of counter-models is a key notion that forms a proper dual to validity --- whereas the satisfiability of the negation is not necessary for invalidity (in contrast to classical logic, where it is a proper dual to validity).
Transforming and reducing counter-models to a normal form is also what ultimately enables our translation.

>>>>

The interest in intuitionistic logic has lead to the development of a number of automated theorem proving systems and a collection of benchmark problems (see e.g. the ILTP library website~\cite{iltp}).
The progress in automated reasoning for intuitionistic logic, however, has been slower than the  impressive advances in solvers for classic logics --- evidenced, e.g., by the CASC~\cite{casc} and SAT~\cite{satc} competitions.
%, e.g., the first-order provers~\cite{kovacs2013first, schulz2002brainiac, korovin2008iprover}
%Comparing benchmarks for classical and intuitionistic automated theorem proving, e.g. TPTP~\cite{casc} and , we see that classical provers currently have a far better success rate. Furthermore all the most popular first-order utilize classical logic.
This difference can partially be explained by fundamental differences between the logics.
First of all, determining intuitionistic validity is computationally harder, e.g., in the propositional case intuitionistic validity is \verb+PSPACE+-complete~\cite{statman1979intuitionistic}, whereas classical validity is \verb+coNP+-complete~\cite{cook1971complexity}.
A further advantage of classical logic is the existence of calculi that are particularly suited for automation, such as superposition~\cite{bachmair2001resolution}, which rely on the existence of convenient normal forms such as CNF, and the duality between validity and satisfiability (i.e., in order to show the validity of a formula it suffices to show the unsatisfiability of the negated formula, which is insufficient in intuitionistic logic).
The first dedicated intuitionistic theorem provers~\cite{mclaughlin2009efficient,tammet1996resolution} used the naive inverse method, i.e., a direct search for a cut-free proof by applying the rules from some proof calculus inversely, which generally leads to a very complex search. 
More recently, connection-based methods have been applied to various non-classical logics~\cite{otten2005clausal,otten2021nanocop}, including intuitionistic logic.
There have also been some successful attempts to study intuitionistic validity via embedding into higher-order classical logic~\cite{LEO}.
We finally add that, in contrast to intuitionistic provers, a tremendous amount of work has been put into optimizing provers for classical logic.

With this work we want to leverage the progress in classical reasoning for intuitionistic logic.
To this end, we give an embedding from intuitionistic logic into classical logic. 
That is, we give an effective procedure that returns for each formula $\varphi$ a modified formula $\varphi^\#$, such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid.
Interestingly, the reverse direction, the famous double-negation translation, has long been established and goes back to Glivenko~\cite{glivenko1929quelques} in the propositional case, and to G\"odel~\cite{godel1933intuitionistischen} and Gentzen~\cite{gentzen1936widerspruchsfreiheit} in the first-order case. In the propositional case it is particularly simple: $\varphi$ is classically valid if and only if $\neg\neg\varphi$ is intuitionistically valid. Intuitively the translation collapses for each subformula $\psi$ of $\varphi$ the truth values of $\psi$ and $\neg\neg\psi$, which are classically but not intuitionistically equivalent. This gives us a first idea why the reverse direction is perhaps more difficult: We need to expand the truth values of $\psi$ and $\neg\neg\psi$, i.e. if they both occur in $\varphi$, we must have a way to (classically) assign different truth values to their respective counterparts in $\varphi^\#$. In particular, this necessitates the introduction of new propositional variables in our procedure, which already marks a big difference to the double negation translation.
%
In addition to the translation of formulas we give an effective translation of counter-models, i.e. for each intuitionistic counter-model of $\varphi$ we can effectively construct a classical counter-model of $\varphi^\#$ and vice versa.
We note that the existence of counter-models is a key notion that forms a proper dual to validity --- whereas the satisfiability of the negation is not necessary for invalidity (in contrast to classical logic, where it is a proper dual to validity).
Transforming and reducing counter-models to a normal form is also what ultimately enables our translation.
